\sys: Automated Data-Parallel Shell Processing

### 2020-05-01 pldi/draft.tex
Update deadlock

\heading{Dangling FIFOs and Zombie Producers}
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
% \ttt{SIGPIPE} 
However, when its consumer exits early, the producer needs to be notified so that it does not keep appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.

In \unix, write errors are communicated differently from errors for other system calls.
Whereas other system calls return an error, write errors are communicated through an out-of-band mechanism (signal handler).
\unix is unusual in this regard from other systems,\footnote{
  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
}
because pipes and pipelines are at the heart of its philosophy.
Unfortunately, if the pipe hasn't been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
Consider the following case:
\begin{lstlisting}[language=sh, numbers=none]
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the example above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves to the second \ttt{cat} unable make process, as it cannot write (and thus process) its output but also unwire that its consumer has exited.
Coupled with \ttt{wait} at the end, it causes the entire pipeline to deadlock.

To solve this problem, \sys emits cleanup logic that operates from the end of the pipeline and towards its start.
The emitted code first gathers the ID of the last process and passes it as a parameter to \ttt{wait}; 
  this causes \ttt{wait} to block only on the last consumer of the pipeline.
Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
% need to define split and merge points; this should be "up to next merge point upstream"

>>>>

\heading{Dangling FIFOs and Zombie Producers}
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a producer commonly exits after it has produced and sent all of its output to a pipe.
% \ttt{SIGPIPE} 
However, if its consumer exits early, the producer needs to be notified so that it stops appending to the stream. % ---so that it avoids unnecessary work and buffering challenges in the kernel.

In \unix, this is achieved by an out-of-band error mechanism:
  the operating system will deliver a \ttt{PIPE} signal to the producer, notifying it that the pipe's consumer has exited.
This is different from errors for other system calls return an error and unusual compared to non-\unix systems\footnote{
  For example, Windows indicate errors for \ttt{WriteFile} by its return code---similar to \ttt{DeleteFile} and other Win32 functions.
}
primarily because pipes and pipelines are at the heart of its philosophy.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet---which is possible for named pipes---\unix cannot signal this condition.
Consider the following emitted code from \sys's earlier versions:
\begin{lstlisting}[language=sh, numbers=none]
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{lstlisting}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves the second \ttt{cat} unable to make process, both (i) blocked unable to write output, and (ii) unaware that its consumer has exited.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

To solve this problem, \sys emits cleanup logic that operates from the end of the pipeline and towards its start.
The emitted code first gathers the ID of the last process and passes it as a parameter to \ttt{wait}; 
  this causes \ttt{wait} to block only on the last consumer of the pipeline.
Right after \ttt{wait}, \sys inserts a routine that delivers \ttt{PIPE} signals to any remaining processes upstream.
This epilogue is repeated for every \todo{merge/join} point in the transformed pipeline.
% Earlier: need to define split (fork?) and merge (join?) points; this should be "up to next merge point upstream"
